---
title: "Chapter 8. Restricted Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Chapter 8. Restricted Estimation"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Load the data and create variables

We process the data with the following code:

```{r}
library(hansen)

xx <- as.matrix(cbind(
  1,
  log(mrw1992$y60),
  log(mrw1992$invest / 100),
  log(mrw1992$pop_growth / 100 + 0.05),
  log(mrw1992$school / 100)
))

lndY <- log(mrw1992$y85) - log(mrw1992$y60)

y <- as.matrix(lndY[mrw1992$n == "Yes"])
x <- xx[mrw1992$n == "Yes", ]
```

# Table 8.1

Using the OLS functions from Chapter 4, we can compute the slopes and standard
errors with the following C++ code:

```cpp
Mat<double> ols3_(const doubles_matrix<>& y, const doubles_matrix<>& x) {
  Mat<double> Y = as_Mat(y);
  Mat<double> X = as_Mat(x);

  int n = x.nrow();
  int k = x.ncol();

  Mat<double> beta_ols = ols_(Y, X);
  Mat<double> e_ols = Y - (X * beta_ols);
  Mat<double> Xe_ols = X.each_col() % e_ols;
  Mat<double> invX = inv(X.t() * X);
  Mat<double> V_ols = (n / (n - k)) * invX * (Xe_ols.t() * Xe_ols) * invX;
  Mat<double> se_ols = sqrt(V_ols.diag());

  Mat<double> res = join_rows(beta_ols, se_ols);
  return res;
}

[[cpp11::register]] doubles_matrix<> ols3_r_(const doubles_matrix<>& y,
                                             const doubles_matrix<>& x) {
  return as_doubles_matrix(ols3_(y, x));
}
```

Now we can use the function in R:

```{r}
table81_1 <- ols3(y, x)

rownames(table81_1) <- c(
  "Intercept", "log(GDP 1960)", "log(1/GDP)",
  "log(n+g+delta)", "log(School)"
)
colnames(table81_1) <- c("Estimate", "Std. Error")

round(table81_1, 3)
```

Now we create a new function for the constrained regression in C++:

```cpp
Mat<double> cns_(const doubles_matrix<>& y, const doubles_matrix<>& x,
                 const doubles_matrix<>& r) {
  Mat<double> Y = as_Mat(y);
  Mat<double> X = as_Mat(x);
  Mat<double> R = as_Mat(r);

  int n = x.nrow();
  int k = x.ncol();

  Mat<double> invX = inv(X.t() * X);
  Mat<double> iR = invX * R * inv(R.t() * invX * R) * R.t();
  Mat<double> beta_ols = ols_(Y, X);
  Mat<double> beta_cns = beta_ols - (iR * beta_ols);
  Mat<double> e_cns = Y - (X * beta_cns);
  Mat<double> Xe_cns = X.each_col() % e_cns;
  Mat<double> V_tilde = (n / (n - k + 1)) * invX * (Xe_cns.t() * Xe_cns) * invX;
  Mat<double> V_cls = V_tilde - (iR * V_tilde) - (V_tilde * iR.t()) +
    (iR * V_tilde * iR.t());
  Mat<double> se_cns = sqrt(V_cls.diag());

  Mat<double> res = join_rows(beta_cns, se_cns);
  return res;
}

[[cpp11::register]] doubles_matrix<> cns_r_(const doubles_matrix<>& y,
                                            const doubles_matrix<>& x,
                                            const doubles_matrix<>& r) {
  return as_doubles_matrix(cns_(y, x, r));
}
```

Now we can use the function in R:

```{r}
r <- as.matrix(c(0, 0, 1, 1, 1))

table81_2 <- cns(y, x, r)

rownames(table81_2) <- rownames(table81_1)
colnames(table81_2) <- colnames(table81_1)

round(table81_2, 3)
```

Now we create a new function for the efficient minimum distance regression in
C++:

```cpp
Mat<double> emd_(const doubles_matrix<>& y, const doubles_matrix<>& x,
                 const doubles_matrix<>& r) {
  Mat<double> Y = as_Mat(y);
  Mat<double> X = as_Mat(x);
  Mat<double> R = as_Mat(r);

  int n = x.nrow();
  int k = x.ncol();

  Mat<double> invX = inv(X.t() * X);
  Mat<double> beta_ols = ols_(Y, X);
  Mat<double> e_ols = Y - (X * beta_ols);
  Mat<double> Xe_ols = X.each_col() % e_ols;
  Mat<double> V_ols = (n / (n - k)) * invX * (Xe_ols.t() * Xe_ols) * invX;

  Mat<double> V_r = V_ols * R * inv(R.t() * V_ols * R) * R.t();
  Mat<double> beta_emd = beta_ols - V_r * beta_ols;
  Mat<double> e_emd = Y - (X * beta_emd);
  Mat<double> Xe_emd = X.each_col() % e_emd;

  Mat<double> V2 = (n / (n - k + 1)) * invX * (Xe_emd.t() * Xe_emd) * invX;
  Mat<double> V_emd = V2 - V2 * R * inv(R.t() * V2 * R) * R.t() * V2;
  Mat<double> se_emd = sqrt(V_emd.diag());

  Mat<double> res = join_rows(beta_emd, se_emd);
  return res;
}

[[cpp11::register]] doubles_matrix<> emd_r_(const doubles_matrix<>& y,
                                            const doubles_matrix<>& x,
                                            const doubles_matrix<>& r) {
  return as_doubles_matrix(emd_(y, x, r));
}
```

Now we can use the function in R:

```{r}
table81_3 <- emd(y, x, r)

rownames(table81_3) <- rownames(table81_1)
colnames(table81_3) <- colnames(table81_1)

round(table81_3, 3)
```

Putting all the results together:

```{r}
table81 <- cbind(
  paste0(
    round(table81_1[, "Estimate"], 3), " (",
    round(table81_1[, "Std. Error"], 3), ")"
  ),
  paste0(
    round(table81_2[, "Estimate"], 3), " (",
    round(table81_2[, "Std. Error"], 3), ")"
  ),
  paste0(
    round(table81_3[, "Estimate"], 3), " (",
    round(table81_3[, "Std. Error"], 3), ")"
  )
)

rownames(table81) <- rownames(table81_1)
colnames(table81) <- c("OLS", "CNS", "EMD")
```
